---
output: 
  github_document:
    df_print: tibble
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
knitr::opts_knit$set(global.par = TRUE)
```

```{r, include=FALSE}
set.seed(1)
options(width = 100)
par(cex.lab=1.5, cex.axis=1.5, cex.main=1.5, cex.sub=1.5)
```

# proDA

<!-- badges: start -->
<!-- badges: end -->

The goal of `proDA` is to identify differentially abundant proteins in label-free
mass spectrometry data. The main challenge of this data are the many missing values.
The missing values don't occur randomly but especially at low intensities. This 
means that they cannot just be ignored. Existing methods have mostly focused on
replacing the missing values with some reasonable number ("imputation") and then
run classical methods. But imputation is problematic because it obscures the
amount of available information. Which in turn can lead to over-confident 
predictions.

`proDA` on the other hand does not impute missing values, but constructs a 
probabilistic dropout model. For each sample it fits a sigmoidal dropout 
curve. This information can then be used to infer means across samples and the
associated uncertainty, without the intermediate imputation step. `proDA`
supports full linear models with variance and location moderation.

## Installation

You can install the development version from [GitHub](https://github.com/const-ae/proDA) with:

``` r
# install.packages("devtools")
devtools::install_github("const-ae/proDA")
```

The pkgdown documentation is available on https://const-ae.github.io/proDA/reference


In the following section, I will give a very brief overview on the main functionality
of the `proDA` package. New users are advised to directly go to section two,
where I give a complete walkthrough and explain in detail, what steps are necessary
for the analysis of label-free mass spectrometry data.

## Quickstart

The three steps that are necessary to analyze the data are

1. Load the data (see vignette on loading MaxQuant output files)
2. Fit the probabilistic dropout model (`proDA()`)
3. Test in which proteins the coefficients of the model differ (`test_diff()`)

```{r quickstart}
# Load the package
library(proDA)
# Generate some dataset with known structure
syn_dataset <- generate_synthetic_data(n_proteins = 100, n_conditions = 2)

# The abundance matrix
syn_dataset$Y[1:5, ]

# The labelling of columns to two conditions
syn_dataset$groups

# Fit the probabilistic dropout model
fit <- proDA(syn_dataset$Y, design = syn_dataset$groups)

# Identify which proteins differ between Condition 1 and 2
test_diff(fit, `Condition_1` - `Condition_2`, sort_by = "pval", n_max = 5)
```

Other helpful functions for quality control are `median_normalization()` and 
`dist_approx()`.


## proDA Walkthrough

`proDA` is an R package that implements a powerful probabilistic dropout model
to identify differentially abundant proteins. The package was designed 
specifically with label-free mass spectrometry data in mind, which poses the
challenge of many missing values.

But all this is useless if you cannot load your data. So in the next section, I 
will give an example how to load the abundance matrix.

### Loading Data

I will now demonstrate how to load a MaxQuant output file. For more information
about different approaches, you can also take a look at the vignette on loading
data.

MaxQuant is one of the most popular tools for handling raw MS data. It produces
a number of files. The important file that contains the protein intensities is
called `proteinGroups.txt`. It is a large table with detailed information about
the identification and quantification process for each protein group (which I will
from now on just call "protein").

This package comes with an example `proteinGroups.txt` file, located in the 
package folder. Naturally, your path will differ because of the installation 
location.

```{r}
system.file("extdata/proteinGroups.txt", package = "proDA", mustWork = TRUE)
```

In this example, I will use the base R functions to load the data, because 
they don't require any additional installed packages.

```{r}
# Load the table into memory
maxquant_protein_table <- read.delim(
    system.file("extdata/proteinGroups.txt", package = "proDD", mustWork = TRUE),
    stringsAsFactors = FALSE
)
```

As I have mentioned, the table contains a lot of information (359 columns!!), but we
are only interested in the columns which contain the measured intensities.

```{r}
# I use a regular expression (regex) to select the intensity columns
intensity_colnames <- colnames(maxquant_protein_table)[grepl("^LFQ\\.intensity\\.",
                                                             colnames(maxquant_protein_table))]
head(intensity_colnames)


# Create the intensity matrix
abundance_matrix <- as.matrix(maxquant_protein_table[, intensity_colnames])
# Adapt column and row maxquant_protein_table
colnames(abundance_matrix) <- sub("^LFQ\\.intensity\\.", "", intensity_colnames)
rownames(abundance_matrix) <- maxquant_protein_table$Protein.IDs
# Print top matrix
abundance_matrix[46:48, 1:6]
```

After extracting the bits from the table we most care about, we will have to adapt the matrix 
slightly. 

First, MaxQuant codes missing values as `0`. This is misleading, because the actual
abundance probably was not zero, but just some value too small to be detected by the mass spectrometer.
Accordingly, I will replace all `0` with `NA`. 

Second, the raw intensity values have a linear mean-variance relation. This is undesirable, because 
a change of `x` units can be a large shift if the mean is small or irrelevant if the mean is large.
Luckily, to make the mean and variance independent, we can just `log` the intensities. Now a change
of `x` units is as significant for highly abundant proteins, as it is for low abundant ones.

```{r}
abundance_matrix[abundance_matrix == 0] <- NA
abundance_matrix <- log2(abundance_matrix)
abundance_matrix[46:48, 1:6]

```




### Quality Control

It is well established truth that quality control (QC) is essential for a successful bioinformatics
project, because mistakes happen. Often we start with normalizing the data to remove potential
sample specific effects. But already this step is challenging, because the missing values cannot
easily be corrected for. Thus, a first helpful plot is to look how many missing values 
each sample contains. 

```{r qc-mis_barplot,  out.width="60%", fig.height=4, fig.align="center"}

barplot(colSums(is.na(abundance_matrix)),
        ylab = "# missing values",
        xlab = "Sample 1 to 36")
```

We can see that the number of missing values differs substantially between samples (between 30% and
90%). If we take a look at the intensity distribution for each sample, we see that they differ
substantially as well. 

```{r qc-raw_boxplot, out.width="60%", fig.height=4, fig.align="center"}
boxplot(abundance_matrix,
        ylab = "Intensity Distribution",
        xlab = "Sample 1 to 36")
```

Note that, the intensity distribution is shifted upwards for samples (for example the last one) 
which also have a large number of missing values. This agrees with our idea that small values are
more likely to be missing. On the other hand, this also demonstrates why normalization methods
such as quantile normalization are problematic. They distort the data until all the distributions
are equal. I will apply the more "conservative" median normalization, which just ignores the 
missing values. It transforms the values so that the median difference between the sample and 
average across all other samples is zero.

```{r}
normalized_abundance_matrix <- median_normalization(abundance_matrix)
```



```{r}
da <- dist_approx(normalized_abundance_matrix)
```


```{r sample_dist, out.width="60%", fig.align="center"}
# This chunk only works if pheatmap is installed
# install.packages("pheatmap")
sel <- c(1:3,  # CG1407
         7:9,  # CG59163
         22:24)# Cg6618

plot_mat <- as.matrix(da$mean)[sel, sel]
plot_mat[diag(9) == 1] <- NA
uncertainty <- matrix(paste0(" ± ",round(as.matrix(da$sd * 1.96)[sel, sel], 1)), nrow=9)
pheatmap::pheatmap(plot_mat, 
                   cluster_rows = FALSE, cluster_cols = FALSE,
                   display_numbers= uncertainty,
                   number_color = "black")
```



### Fitting the Probabilistic Dropout Model

In the next step we will fit the actual linear probabilistic dropoout model to the normalized
data. But before we start I will create a data.frame that contains some additional info on each
sample, in particular to which condition that sample belongs.

```{r}

sample_info_df <- data.frame(name = colnames(normalized_abundance_matrix),
                             stringsAsFactors = FALSE)
sample_info_df$condition <- substr(sample_info_df$name, 1, nchar(sample_info_df$name)  - 3)
sample_info_df$replicate <- as.numeric(
  substr(sample_info_df$name, nchar(sample_info_df$name)  - 1, 20)
)
sample_info_df
```


Now we can call the `proDA()` function to actually fit the model. We specify the `design` using
the formula notation, referencing the `condition` column in the `sample_info_df` data.frame that
we have just created. In addition, I specify that I want to use the `S2R` condition as the reference
because I know that it was the negative control and this way automatically all coefficients
measure the difference between the condition and the negative control.

```{r}
fit <- proDA(normalized_abundance_matrix, design = ~ condition, 
             col_data = sample_info_df, reference_level = "S2R")
fit
```

The `proDAFit` object prints a number of useful information about the convergence of the model,
the size of the dataset, the number of missing values, and the inferred hyper parameters.

To make it easy to find available methods on the `proDAFit` object, the `$`-operator is overloaded
and shows a list of possible functions:

```{r}
# Equivalent to feature_parameters(fit)
fit$feature_parameters
```

Internally `proDAFit` object is implemented as a subclass of `SummarizedExperiment`.
This means it can be subsetted, which can be particularly useful for calculating the distance 
of a subset of proteins and samples.

```{r protein_dist, out.width="60%", fig.align="center"}
# This chunk only works if pheatmap is installed
# install.packages("pheatmap")
pheatmap::pheatmap(dist_approx(fit[1:20, 1:3], by_sample = FALSE)$mean)
```



### Differential Abundance

Finally we can identify

```{r}
test_res <- test_diff(fit, "conditionCG1407")
test_res
```






